### 2015-04-15 Dissertation/chapter-benchmarking.tex
Copied Comprehensive to Disseration
Major update from journal article polishing

\chapter{Benchmarking P3}
\label{chap-benchmarking}
TODO Explain the purpose of this chapter

@@

Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
are treated as requiring more evaluations than any successful run. If the median run was not successful
no point is shown. The maximum problem size used for each problem was set to be the largest size we could
feasibly determine LTGA's optimal population size. For many larger problems results are not shown
for hBOA due to the extreme computational cost required to optimally set the population size.

\subsection{Quantitative Comparison}
Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.

\subsection{Local Search}
The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
sizes. This is especially true for the three randomly generated problem problem classes. These problems
may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
better than any other algorithm. We believe this is
because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
flipping entire traps.
This ability also leads $1+(\lambda, \lambda)$
to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.

@@

@@

Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
In order to calculate this value we recorded how many times search rebuilt the model during each run.
Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
no more than twice the algorithmic cost of an evaluation even using 2048 bits.

When applying a crossover subset, P3 tries random donors from the population until one is found with at least
one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
important note is that each donation may range in size from a single bit up to $N-1$, with smaller
donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
in P3, but its unlikely to be very high.

@@

To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
of 100 runs, with unsuccessful runs treated as slower than successful runs.
In total these runs
took over 6 computing years to complete. We were therefore forced to use
diverse hardware potentially running multiple processes in parallel.
As such these timings should be considered very skeptically as a number
of sources may contribute to biased error in the results. However, due to
the size of the dataset and the general regularity in the results some general conclusions
can still be made.

\subsubsection{Model Building}
hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
As P3 and LTGA require a similar asymptotic complexity per
evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.

\subsubsection{P3}
Any increase in P3's algorithmic complexity is offset by its efficient use of evaluations.
In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.
On
Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT P3 still finds the global
optimum faster than all comparison techniques. LTGA is able to match P3's performance almost exactly on HIFF,
and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
than LTGA but by no more than a constant factor.

>>>>

\chapter{Benchmarking P3}
\label{chap-benchmarking}

@@

Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
algorithms to find the global optimum for multiple sizes of each problem. Each data point in
Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
are treated as requiring more evaluations than any successful run. If the median run was not successful
no point is shown.
Medians are used as the data is not normally distributed, and because it allows for more meaningful comparison
between techniques with different success rates.
The maximum problem size used for each problem was set to be the largest size we could
feasibly determine LTGA's optimal population size. For many larger problems results are not shown
for hBOA due to the extreme computational cost required to optimally set the population size.

\subsection{Quantitative Comparison}
Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
The likelihood that P3 would achieve these pairwise results assuming its median result is actually worse is $p < 10^{-15}$
according to the binomial test.
Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
$p < 10^{-5}$ for all problems.

\subsection{Local Search}
The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
sizes. This is especially true for the three randomly generated problem classes. These problems
may contain relatively few local optima or just be exceptionally difficult for the model based algorithms.
On Deceptive Trap and Deceptive Step Trap using 4 or less traps, $1+(\lambda, \lambda)$ performs significantly
better than any other algorithm. We believe this is
because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
flipping entire traps.
This ability also leads $1+(\lambda, \lambda)$
to outperform the Random Restart Hill Climber on all problems except Nearest Neighbor NK.

@@

At 58\% of the recording intervals hBOA has the worst fitness of any solver. Most of the exceptions
occur when hBOA is still evaluating its initial population, allowing this random search to temporarily surpass the
local search methods. After $N$ evaluations, however, hBOA and LTGA both fall behind until their models begin
to improve.
Parameter-less hBOA reaches intermediate fitnesses faster than hBOA, doing so on 62\% of intervals, as its models begin
to optimize earlier than hBOA.  However, this trend is reversed after a sufficient number of evaluations, most clearly
on Deceptive Step Trap and Ising Spin Glasses, as hBOA's tuned population overtakes Parameter-less hBOA's parallel populations.

@@

Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
In order to calculate this value we recorded how many times search rebuilt the model during each run.
Figure~\ref{fig-rebuilds} shows the estimated ratio of model rebuilding cost ($N^2$ per rebuild) over evaluation cost ($N$ per evaluation).
If the cost of model building
scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this is the case. For both Trap problems
and HIFF there is slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
no more than twice the algorithmic cost of an evaluation even using 2048 bits.

When applying a crossover subset, P3 tries random donors from the population until one is found with at least
one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
important note is that each donation may range in size from a single bit up to $N-1$. However,
repeated donation attempts are far more likely to happen with smaller clusters.
As such this may cause some super-linear growth
in P3, but its unlikely to be very high.

@@

To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}.
Similar to Figure~\ref{fig-evals-to-success}, each point represents the median
of 100 runs, with unsuccessful runs treated as slower than successful runs.
\begin{comment}
In total these runs
took over 6 computing years to complete. We were therefore forced to use
diverse hardware potentially running multiple processes in parallel.
As such these timings should be considered very skeptically as a number
of sources may contribute to biased error in the results. However, due to
the size of the dataset and the general regularity in the results some general conclusions
can still be made.
\end{comment}
These results were collected using 2.5GHz Intel Xeon E5-2670v2 processors.

\subsubsection{Model Building}
hBOA and Parameter-less hBOA perform much worse when using wall clock time as the unit of comparison than when using
evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
As P3 and LTGA require a similar asymptotic complexity per
evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.

\subsubsection{P3}
When LTGA is optimally tuned to a single instance problem with an efficient evaluation function
it can find the global optimum faster than P3 in terms of wall clock time. However, on randomly
generated problem classes P3's efficient use of evaluations is enough to overtake LTGA.

On the four single instance problems LTGA not only finds the global optimum using less wall clock time,
the factor speedup increases as problem length does. Na\"{i}vely this suggests LTGA is achieving
a lower order of complexity. However, for these experiments LTGA is growing at sub-linear time per
evaluation, which is not asymptotically stable due to (at minimum) the time required to perform an evaluation.
We suspect that the true cause is that $N$ is small enough to be overshadowed by lower order polynomial terms.
For example, LTGA requires $O(N/\mu)$ time per evaluation to rebuild the linkage
model from the frequency table. As a result, for small $\mu$ model building, and not extracting pairwise
frequency, can dominate runtime.

When applied to randomly generated problem classes, the differences in P3 and LTGA's evaluation complexity
dominates runtime complexity. Similar to with Figure~\ref{fig-evals-to-success}, the amount of speedup P3
achieves over LTGA increases with problem size on Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT.

Across both types of problems we find that P3's time per evaluation grows approximately linearly. As such,
we conclude that P3 requires asymptotically similar amounts of time per evaluation as the other efficient techniques.
